### 2023-03-27 doc/_bayes_meanfit.qmd
finish writing model section

The real advantage of rewriting the joint distribution as in
@eq-joint_freq_gamma comes from splitting this joint distribution as a product
of conditional distributions of the form
$$
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_to_prod_gamma}
Written in this form, we can finally propose a probabilistic model for how the
mean fitness relates to the frequency ratios we determine in our experiments.
The second term on the right-hand side of @eq-joint_to_prod_gamma relates how
the determined frequency ratio $\gamma_t^{(b)}$ relates to the mean fitness
$\bar{s}_t$. From @eq-fitness_ratio_neutral and @eq-gamma_def, we can write
$$
\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
$${#eq-log_ratio_neutral}
where we added an extra term, $\varepsilon_t^{(n)}$, characterizing the 
deviations of the measurements from the theoretical model. It is common to 
assume that these errors are normally distributed with mean zero and some 
standard deviation $\sigma_t$, implying that
$$
\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
$${#eq-log_gamma_normal}
where we include the nuisance parameter $\sigma_t$ to be determined. If we 
assume the log frequency ratio is normally distributed, this implies the 
frequency ratio itself is distributed log-normal. This means that
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
$${#eq-meanfit_likelihood}
Having added the nuisance parameter $\sigma_t$ implies that we must update
@eq-meanfit_bayes to
$$
\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
\pi(\bar{s}_t) \pi(\sigma_t),
$${#eq-meanfit_bayes_full}
where we assume the prior for each parameter is independent, i.e.,
$$
\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
$${#eq-meanfit_indep_prior}

For the first term on the right-hand side of @eq-joint_to_prod_gamma,
$\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)$, we remove the conditioning
on the population mean fitness since it does not add any information on top of
what the frequency ratio $\gamma_t^{(n)}$ already gives. Therefore, we have
$$
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
$${#eq-freq_cond_gamma}
The right-hand side of @eq-freq_cond_gamma asks us to compute the probability
of observing a frequency value $f_t^{(n)}$ given that we get to observe the
ratio $\gamma_{t}^{(n)}$. If the ratio happened to be $\gamma_{t}^{(n)} = 2$,
we could have $f_{t+1}^{(n)} = 1$ and $f_{t+1}^{(n)} = 0.5$, for example.
Although, it would be equally likely that $f_{t+1}^{(n)} = 0.6$ and
$f_{t+1}^{(n)} = 0.3$ or $f_{t+1}^{(n)} = 0.1$ and $f_{t+1}^{(n)} = 0.05$ for
that matter. As a matter of fact, given that we only get to observe the
frequency ratio $\gamma_t^{(n)}$, we can only restrict the values of the
frequency $f_{t}^{(n)}$ to be
$$
f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
$${#eq-freq_given_gamma_range}
A priori, we do not have any reason to favor any value over any other, therefore
it is natural to write
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
$${#eq-f_given_gamma_uniform}

@@

Putting all the pieces we have developed in this section together results in an
inference for the population mean fitness values of the form
$$
\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
$${#eq-meanfit_full_inference}
where we have
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
$${#eq-meanfit_eq1}
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
$${#eq-meanfit_eq2}
$$
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
$$
and
$$
sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
$$
where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
choose.

>>>>

The real advantage of rewriting the joint distribution as in
@eq-joint_freq_gamma comes from splitting this joint distribution as a product
of conditional distributions of the form
$$
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)
\pi(\gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_to_prod_gamma}
Written in this form, we can finally propose a probabilistic model for how the
mean fitness relates to the frequency ratios we determine in our experiments.
The second term on the right-hand side of @eq-joint_to_prod_gamma relates how
the determined frequency ratio $\gamma_t^{(b)}$ relates to the mean fitness
$\bar{s}_t$. From @eq-fitness_ratio_neutral and @eq-gamma_def, we can write
$$
\ln \gamma_t^{(n)} = - \bar{s}_t + \varepsilon_t^{(n)},
$${#eq-log_ratio_neutral}
where, for simplicity, we set $\tau = 1$. Note that we added an extra term,
$\varepsilon_t^{(n)}$, characterizing the deviations of the measurements from
the theoretical model. It is common to assume that these errors are normally
distributed with mean zero and some standard deviation $\sigma_t$, implying that
$$
\ln \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\mathcal{N}\left(-\bar{s}_t, \sigma_t \right),
$${#eq-log_gamma_normal}
where we include the nuisance parameter $\sigma_t$ to be determined. If we 
assume the log frequency ratio is normally distributed, this implies the 
frequency ratio itself is distributed log-normal. This means that
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t  \sim 
\log \mathcal{N}\left(-\bar{s}_t, \sigma_t \right).
$${#eq-meanfit_likelihood}
Having added the nuisance parameter $\sigma_t$ implies that we must update
@eq-meanfit_bayes to
$$
\pi(\bar{s}_t, \sigma_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, \gamma_t^{(n)} \mid \bar{s}_t, \sigma_t) 
\pi(\bar{s}_t) \pi(\sigma_t),
$${#eq-meanfit_bayes_full}
where we assume the prior for each parameter is independent, i.e.,
$$
\pi(\bar{s}_t, \sigma_t) = \pi(\bar{s}_t) \pi(\sigma_t).
$${#eq-meanfit_indep_prior}
For numerical stability, we will select weakly-informative priors for both of
these parameters. In the case of the nuisance parameter $\sigma_t$, the prior
must be restricted to positive values only, since standard deviations cannot be
negative.

For the first term on the right-hand side of @eq-joint_to_prod_gamma,
$\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t)$, we remove the conditioning
on the population mean fitness since it does not add any information on top of
what the frequency ratio $\gamma_t^{(n)}$ already gives. Therefore, we have
$$
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}, \bar{s}_t) =
\pi(f_t^{(n)} \mid \gamma_{t}^{(n)}).
$${#eq-freq_cond_gamma}
The right-hand side of @eq-freq_cond_gamma asks us to compute the probability
of observing a frequency value $f_t^{(n)}$ given that we get to observe the
ratio $\gamma_{t}^{(n)}$. If the ratio happened to be $\gamma_{t}^{(n)} = 2$,
we could have $f_{t+1}^{(n)} = 1$ and $f_{t+1}^{(n)} = 0.5$, for example.
Although, it would be equally likely that $f_{t+1}^{(n)} = 0.6$ and
$f_{t+1}^{(n)} = 0.3$ or $f_{t+1}^{(n)} = 0.1$ and $f_{t+1}^{(n)} = 0.05$ for
that matter. If we only get to observe the frequency ratio $\gamma_t^{(n)}$, we
know that the numerator $f_{t+1}^{(n)}$ can only take values between zero and
one, all of them being equally likely given only the information on the ratio.
As a consequence, the value of the frequency in the denominator $f_{t}^{(n)}$ is
restricted to fall in the range
$$
f_{t}^{(n)} \in \left(0, \frac{1}{\gamma_t^{(n)}} \right].
$${#eq-freq_given_gamma_range}
A priori, we do not have any reason to favor any value over any other, therefore
it is natural to write
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform}\left( 0, \frac{1}{\gamma_t^{(n)}} \right).
$${#eq-f_given_gamma_uniform}

@@

Putting all the pieces we have developed in this section together results in an
inference for the population mean fitness values of the form
$$
\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
$${#eq-meanfit_full_inference}
where we have
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
$${#eq-meanfit_eq1}
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
$${#eq-meanfit_eq2}
$$
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
$${#eq-meanfit_eq3}
and
$$
\sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
$${#eq-meanfit_eq4}
where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
choose.
